---
title: "Lorelogram: describing correlation structure in binary data"
date: "`r format(Sys.time(), '%d %B %Y')`"
bibliography: C:/Users/Fabiola/Documents/Bibtex/Lorelogram Vignette.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lorelogram: describing correlation structure in binary data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Table of contents
1. [Introduction](#introduction)
2. [How to apply the lorelogram](#paragraph1)
    1. [Data format](#subparagraph1)
3. [Another paragraph](#paragraph2)




# Introduction <a name="introduction"></a>

The __lorelogram__ package provides functions to explore and graphically describe correlation structures in binary data. The package is available on GitHub and can be downloaded in R (@RCoreTeam2018) via:

```{r load pkg}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("FabiolaIannarilli/lorelogram", auth_token = "69451dbc8b643e18d5e25b488e0a485b59a93512") 

```


The method has been developed by @Heagerty1998 and is based on pairwise log-odds ratio. For more information on the method and on applications in ecological studies see @Iannarilli.

Here we focused on guiding new users on how to apply the lorelogram to their own dataset. We provide step-by-step examples, and a comparison of the present function with other functions for calculating the lorelogram in R (@RCoreTeam2018). 

The main function in the __lorelogram__ package is called _lorelogram_. This function allows users to calculate log-odds ratio at each time lag up to a maximum lag between observations in the same sampling unit. It also return a plot of these values, the lorelogram plot, that graphically represents how the correlation change with the increase of the spatial or temporal lag between observations. 

We use the data associated with the package to provide a first example. This dataset consists of detection/nondetection data of gray fox (_Urocyon cinereoargenteus_) collected using camera traps at 100 locations in Northern Minnesota. We provided further details on the structure of this data in the [Data format] section. Here we just load the data and apply the _lorelogram_ function. 

```{r ex1}
data("GrayFox_Hour")
lorelogram(GrayFox_Hour)
```

# How to apply the lorelogram


## Data format <a name="data format"></a>






# References
